High geomagnetic field intensity recorded by anorthosite xenoliths requires a vigorous late Mesoproterozoic geodynamo

### 2022-01-13 PNAS_manuscript/PNAS_manuscript.tex
edit through Cooling rate correction and secular variations section

% Please add a significance statement to explain the relevance of your work
\significancestatement{Acquiring high-fidelity observational records of ancient magnetic field intensity from remanent magnetization of rocks is crucial for constraining the evolution of Earth's core. However, robust estimates of ancient field strengths are often difficult to recover due to alterations or non-ideal rock magnetic behaviors. In this study, we use plagioclase cumulates that formed in the deep crust and were brought to the near surface by magma where they cooled and acquired thermal remanent magnetizations. These plagioclase from the North American Midcontinent Rift have experienced minimal post-formation alterations and record a present-day-like virtual dipole moment 1.1 billion years ago. The results indicate the existence of strong power sources for Earth's dynamo during the late Mesoproterozoic. }

@@

\begin{abstract}
Obtaining estimates of Earth's magnetic field strength in deep time is complicated by non-ideal rock magnetic behavior in many igneous rocks. In this study, we target \textit{ca.} 1092 Ma anorthosite xenoliths that cooled and acquired their magnetization within Beaver River diabase intrusions of the North American Midcontinent Rift. In contrast to the diabase which has non-ideal paleointensity behaviors and fails to provide reliable paleointensity estimates, the anorthosite xenoliths are unusually high-fidelity recorders yielding high-quality, single-slope paleointensity results that are consistent at specimen and site levels. An average value of $\sim$89 ZAm$^2$ for the virtual dipole moment from the anorthosites is higher than that of the dipole component of Earth's magnetic field today. Such a strong late Mesoproterozoic dipole recorded by the anorthosites requires the existence of a vigorous geodynamo at the time. Together with previous paleointensity data from other Midcontinent Rift rocks, our new results indicate that a dynamo with strong power sources persisted through $\sim$25 Myr at \textit{ca.} 1.1 Ga. These observations are inconsistent with there being a progressive monotonic decay of Earth's dynamo strength through the Proterozoic Eon.
\end{abstract}

@@

Paleomagnetic records from ancient rocks are one of the few types of observational data that have the potential to record the thermal evolution of Earth’s core. Evidence for a persistent magnetic field through the Proterozoic, for example, necessitates the existence of plate tectonics that sustained core-mantle boundary heat flow. However, strikingly low estimates of geomagnetic field strengths have been obtained in the Ediacaran Period \cite{Bono2019a, Thallner2021b}, potentially indicating an unusual period of core dynamo regime at the time.

% Those paleointensity data has been interpreted to be consistent with thermal evolution models that use high core thermal conductivity values, where the nucleation of the inner core did not begin to provide compositional convection power for driving the geodynamo until after \textit{ca.} 565 Ma, suggesting that the Proterozoic dynamo might have experienced a monotonic decay when compositional convection was lacking and thermal convection was progressively decreasing \cite{Driscoll2016a, Davies2021a}.

Determinations of ancient magnetic field strengths rely on igneous rocks that acquire thermal remanent magnetizations as they cool in Earth's magnetic field. These magnetizations need to be unmodified by subsequent heating or chemical alterations in order to be able to faithfully record ancient geomagnetic field through time. Intraplate magmatic events are therefore an important target as they can be well-preserved within continental interiors at distance from mountain-building events along continental margins. However, magmatic events such as the formation of large igneous provinces that are often associated with intraplate magmatism are typically of geologically short duration with the bulk of magmatic products emplaced within 1 Myr. The Midcontinent Rift is an exception as magmatism lasted $\sim$25 Myr from 1109 Ma to 1084 Ma. Additionally, extension ceased in the Midcontinent Rift prior to lithosphere separation, preserving volcanics, intrusives, and sediments of the rift within the continental interior. As a result, rocks of the rift have simple paleomagnetic behavior and form a central record of Mesoproterozoic paleogeography \cite{Swanson-Hysell2021c}. The simple magnetizations held by igneous rocks that formed through an extended duration from 1109 to 1084 Ma that is longer than the Neogene Period have the potential for a well-preserved window into the intensity of Earth's magnetic field .

Despite the exquisite preservation of the rocks, non-ideal paleointensity behaviors and overall low success rates have often led to high failure rates and challenged the interpretation of many previous paleointensity results from the Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a}. One typical failure of paleointensity experiment is the double-slope behavior, as two distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope: a higher paleointensity estimate from the low-temperature portion and a lower paleointensity estimate from the high-temperature portion. Ref. \citealp{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas ref. \citealp{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Results with such non-ideal behavior were rejected by ref. \citealp{Sprain2018a} who applied stricter paleointensity selection criteria.

In this study, we target a unique rock type - high-purity anorthosite xenoliths. These anorthosite xenoliths were brought to the near surface with magma that formed sills of the Beaver River diabase \cite{Zhang2021b}. They cooled and acquired their magnetizations in conjunction with the host diabase. Paleointensity experiments on the anorthosite xenoliths have a very high success rate, yielding consistent specimen- and site-level paleointensity results. Rock magnetic data reveal that the anorthosite xenoliths have low anisotropy of thermal remanent magnetization (TRM) and can acquire TRM linearly. Magnetic imaging shows the anorthosite specimens have dominant magnetic carriers interstitial to and within plagioclase crystals and they do not display strong preferred orientations. Our single-slope, high quality paleointensity data add to the reliable paleointensity records during the $\sim$25 Myr of Midcontinent Rift magmatic activity. Our new data from the Midcontinent Rift intrusive rocks support previous data from the volcanics that there was a strong late Mesoproterozoic geodynamo.

@@

While magmatic activity within the Midcontinent Rift was protracted, there were volcanism and intrusions characterized by particularly rapid and voluminous emplacement. The \textit{ca.} 1092 Ma Beaver Bay Complex in northern Minnesota punctuates one such period of magmatism during the Midcontinent Rift main stage activity. The magma that formed the 1091.7 $\pm$ 0.2 Ma Beaver River diabase dikes and sills of the Beaver Bay Complex transported numerous anorthosite xenoliths that have short-axis diameters up to 180 meters via wide conduits \cite{Boerboom2004a, Boerboom2006b}. The anorthosite xenoliths are plagioclase cumulates that formed comagmatically with the host diabase in the lower crust - an interpretation confirmed by U-Pb zircon geochronology \cite{Zhang2021b}.

@@

Thermal modeling results and paleomagnetic directional data show the anorthosite xenoliths to have acquired thermal remanent magnetizations while cooling with the Beaver River diabase \cite{Zhang2021b}. Step-wise thermal demagnetizations show the anorthosite xenoliths have dominantly single-component magnetization that often unblock sharply within temperature ranges between 500\textdegree C and 580\textdegree C \cite{Zhang2021b}.

\section*{Petrography and magnetic imaging of anorthosite xenoliths}

Petrographic imaging shows the dominantly monomineralic anorthosite xenoliths of the Beaver River diabase often develop granoblastic texture characterized by equigranular crystals with poor development of petrofabrics (Fig. \ref{fig:Petro_QDM}A,C). In addition, the plagioclase crystals of the anorthosite xenoliths are often free of large inclusions. Magnetic imaging using the quantum diamond microscope (QDM) shows that the dominant magnetic remanence carriers often exist within and sometimes interstitial to plagioclase crystals (Fig. \ref{fig:Petro_QDM}E).

These features of the Beaver River anorthosite xenoliths are distinct from other anorthosites such as the anorthosite xenoliths within the Duluth Complex Anorthositic Series rocks. The plagioclase of the Duluth Complex anorthosites typically develop interlocking textures that display strong igneous foliations (Fig. \ref{fig:Petro_QDM}B) and associated titanomagnetite mineral fabrics characterized by having aligned needles within plagioclase crystals (Fig. \ref{fig:Petro_QDM}D). Magnetic imaging confirms that the titanomagnetite needles within the plagioclase crystals are often strongly magnetic with magnetic moments of individual needles oriented along their long axes (Fig. \ref{fig:Petro_QDM}D,F). Titanomagnetite and ilmenite symplectic intergrowths also coexist with pyroxene and relict olivine as a product of olivine oxidation in the Duluth Complex plagioclase \cite{Miller1990a}.

The lack of symplecites and exsolved titanomagnetite needles within the Beaver River anorthosite xenoliths distinguish them from plagioclase cumulates of the Duluth Complex and other layered mafic complex where anisotropic magnetic mineral fabrics associated with igneous foliation often occur \cite{Scofield1986a, Selkin2000a, Feinberg2006a}.

@@

\begin{figure*}
\centering
\noindent\includegraphics[width=11.4 cm]{Geologic_map.pdf}
\caption{\footnotesize{Simplified geologic map of Lake Superior region showing the distribution of rift and post-rift-related rocks including extrusive, intrusive, and late/post-rift sediments around Lake Superior. Red diamonds mark sites with paleointensity results that passed the selection criteria from this study. Paleomagnetic sites from ref. \citealp{Pesonen1983a} are categorized by lithology. All sites from ref. \citealp{Pesonen1983a, Kulakov2013a, Sprain2018a} are color-coded by their ages.}}
\label{fig:Geologic_map}
\end{figure*}

% Despite that the anorthosite xenoliths of the Beaver River diabase were entrained as xenoliths and emplaced only $\sim$4 Myr after the formation of the Anorthositic Series rocks of the Duluth Complex when massive plagioclase mushes were generated, $^{206}$Pb/$^{238}$U zircon geochronology data and zircon rare earth element diffusion modeling results suggest it is the most probable that the Beaver River diabase anorthosite xenoliths are entrained cumulate enclaves that formed at the time of Beaver Bay Complex magmatism. The anorthosite xenoliths of the \textit{ca.} 1092 Ma Beaver River diabase has distinct mineralogy from that of the Anorthositic Series rocks of the Duluth Complex. The dominantly monomineralic anorthosite xenoliths have an average anorthite content of $\sim$70\% \cite{Morrison1983a, Doyle2016a}, higher than the average of $\sim$60\% of the Anorthositic Series rocks of the Duluth Complex \cite{Miller1990a}. In addition, petrographic imaging support that the plagioclase from the Beaver River anorthosite formed under a magmatic conditions different from the Duluth Complex anorthositic series magma. While the plagioclase in the Duluth Complex Anorthositic Series typically are euhedral with an interlocking texture (Fig. \ref{Petro_QDM}A), the more calcium-rich plagioclase of the Beaver River anorthosite often develop granoblastic texture characterized by equigranular plagioclase crystals (Fig. \ref{Petro_QDM}B). Abundant magnetite and ilmenite needles that are typically 10 $\mu$m long can be found to have preferred orientations within Duluth Complex plagioclase. Interstitial vermicular titanomagnetite and ilmenite symplectic intergrowths also coexist with pyroxene and relict olivine as a product of olivine oxidation in the Duluth Complex Anorthositic Series rocks \cite{Miller1990a}.

\begin{figure*}
\centering
\noindent\includegraphics[width=11.4 cm]{Petro_QDM.png}
\caption{\footnotesize{Thin section petrographic and magnetic images of anorthosites from the Beaver River anorthosite xenolith from which paleomagnetic site AX16 and geochronology sample MS99033 were collected (left column; ref. \citealp{Zhang2021b}) and and distinct anorthosite from the Duluth Complex Anorthositic Series (right column). Cross-polarized petrographic images (A) Beaver River anorthosite have plagioclase with a granoblastic texture; (B) Plagioclase of Duluth Complex anorthosite have euhedral, interlocking crystals showing igneous foliation; (C) zoomed-in image of plagioclase from MS99033 showing the crystals are largely free of large opaque inclusions; (D) Duluth Complex plagioclase can contain abundant Fe-Ti oxide needles that have preferred orientations that are often parallel with the (010) planes of plagioclase. Magnetic field maps acquired by the quantum diamond microscope (QDM) show that the relatively weak magnetic sources within and interstitial to plagioclase crystals exist in the Beaver River anorthosite in (E) while large oxide needles within Duluth Complex plagioclase are strongly magnetic as evidenced through the complex overlapping magnetic fields above the oxides in (F). Magnetic field maps of (E) and (F) share the same spatial and magnetic scale. The unit of the axes are in $\mu$m.}}
\label{fig:Petro_QDM}
\end{figure*}

@@

Following IZZI-modified paleointensity experiments \cite{Yu2004a}, 40 from a total of 86 anorthosite specimens and 0 out of a total of 69 diabase specimens passed our paleointensity result selection criteria of having (1) a maximum angular deviation (MAD; ref. \citealp{Kirschvink1980a}) of $<$20\textdegree; (2) scatter parameter ($\beta$; ref. \citealp{Coe1978a}) values of $<$15 per cent; (3) a deviation angle (DANG; ref. \citealp{Tauxe2004a}) of $<$5\textdegree; (4) fraction of remanence fitted for paleointensity estimate (FRAC; ref. \cite{Shaar2013a}) $>$0.6; (5) scatter statistic (SCAT; ref. \citealp{Shaar2013a}) = TRUE; (6) a maximum magnetic moment difference between adjacent zero-field steps (GAP-Max; ref. \citealp{Shaar2013a}) $<$ 0.25; (7) number of pTRM checks $>$ 2; (8) and number of measurements used for paleointensity determination $\geq$ 4. 7 anorthosite sites and no diabase site-level results pass our selections. Example NRM/TRM plots are shown in Fig. \ref{fig:IZZI_examples}. Summary specimen absolute paleointensity estimates and site-level weighted mean paleointensity values are plotted in Fig. \ref{fig:PINT_cooling_corrected} (and provided in Table S1). The cooling rate-corrected absolute paleointensity estimates from all anorthosite sites have a weighted mean of 52.72 $\pm $5.72 $\mu$T. This corresponds to a calculated virtual dipole moment of $\sim$89 ZAm$^2$ ($10^{21} Am^2$) \textit{ca.} 1092 Ma. All measurement-level paleointensity experiment data is available within the MagIC database (\url{}; UPDATE WHEN DOI IS GENERATED).

\begin{figure*}
\noindent\includegraphics[width=17.8cm]{IZZI_examples.pdf}
\centering
\caption{\footnotesize{Example results of paleointensity experiments displayed on NRM/TRM plots and zero-field heating results are shown on orthogonal plots (insets) for anorthosite and diabase specimens. Red (blue) circles indicate zero-field/in-field (in-field/zero-field) steps `ZI’ (`IZ’). Triangles mark partial thermal remanent (pTRM) checks. Blue and red squares in the Zijderveld plots are X–Y and X–Z projections, respectively, of the NRMs in specimen coordinates. Plots on the top row show successful specimen paleointensity results with straight, single-slope behaviors that pass our selection criteria. The green lines represent our fitting for the dominant single-slope component that passes our acceptance criteria with estimates of the ancient field (B$_{anc}$) being used to determine site means. The plots for anorthosite specimens AX1-1a and diabase BD5-1b on the bottom row show typical non-ideal sagging and zigzagging behaviour that fails our acceptance criteria. Plot for diabase specimen BD2-5a shows typical paleointensity behavior of this site - specimens typically have double-slope behaviors although the lower temperature slopes can pass the selection criteria, resulting in the high estimated B$_{anc}$ values shown in Fig. \ref{fig:PINT_cooling_corrected}. Thus, the estimates of B$_{anc}$ for this specimen plot is illustrative, but were not used to calculate a site mean paleointensity value in Fig. \ref{fig:PINT_cooling_corrected}.}}
\label{fig:IZZI_examples}
\end{figure*}

Paleointensity experiment results from the anorthosite xenoliths and their host diabase show three general types of dominant specimen behaviors on NRM/TRM plots (Fig. \ref{fig:IZZI_examples}): dominantly single-slope, double-slope or sagging, and zigzagging. Typical specimen paleointensity behavior of the anorthosites display straight, single-slope NRM/TRM plots and the accepted fractions of temperature steps span over the origin-trending, primary remanence components (Fig. \ref{fig:IZZI_examples}). We accept all specimen- and site-level absolute paleointensity results from the anorthosite xenoliths that pass the selection. Diabase specimens failed the selection because of double-slope or sagging behavior (fail FRAC selection; see methods), poor pTRM checks, and zigzagging plots (fail SCAT, DANG selection; see methods; Fig. \ref{fig:IZZI_examples}). A 20 mT alternating field treatment after in-field heating steps was applied to some specimens but it did not result in significant changes in paleointensity behaviors of the anorthosite xenoliths or the diabase (Fig. \ref{fig:PINT_cooling_corrected}; Supporting Information).

In addition to estimating paleointensities by introducing a set of selection criteria to filter our experiment results, we apply an independent method of ref. \citealp{Cych2021a} which uses all experimental data regardless of their NRM/TRM plot statistics to perform bias corrected estimation of paleointensity (BiCEP). This Bayesian probabilistic method is based on an assumption that paleointensity estimates from specimens that come from a same cooling unit are distributed around a true paleointensity value with the various deflections being expressed as the curvature parameter of the NRM/TRM plot \cite{Paterson2011a}. The posterior paleointensity distributions for sites with high-quality specimen-level data are in agreement with site-level weighted-mean results (Fig. \ref{fig:PINT_cooling_corrected}; Fig. S1).

Overall, the high-quality experiment results from the anorthosite xenoliths of the Beaver River diabase and the consistent paleointensity estimates based on independent methods confirm that the anorthosite xenoliths record a high geomagnetic field \textit{ca.} 1092 Ma. The paleointensity quality index (Q$_{PI}$; ref. \citealp{Biggin2014a}) for the anorthosite xenoliths are all above 4 (Table S2).

@@

\subsection*{Rock magnetic experiment}

>>>>

% Please add a significance statement to explain the relevance of your work
\significancestatement{Acquiring high-fidelity observational records of ancient magnetic field intensity from the remanent magnetization of rocks is crucial for constraining the evolution of Earth's core. However, robust estimates of ancient field strengths are often difficult to recover due to alteration or non-ideal rock magnetic behavior. In this study, we use plagioclase cumulates (anorthosite) that formed in the deep crust and were brought to the near surface by magma where they cooled and acquired thermal remanent magnetizations. These anorthosite xenoliths from the North American Midcontinent Rift have experienced minimal alteration since their emplacement and yield high quality paleointensity estimates. These data record a strong present-day-like virtual dipole moment 1.1 billion years ago. A strong field persisted over a 25 million year or greater interval indicating the existence of appreciable power sources for Earth's dynamo during the late Mesoproterozoic. }

@@

\begin{abstract}
Obtaining estimates of Earth's magnetic field strength in deep time is complicated by non-ideal rock magnetic behavior in many igneous rocks. In this study, we target anorthosite xenoliths that cooled and acquired their magnetization within \textit{ca.} 1092 Ma shallowly emplaced diabase intrusions of the North American Midcontinent Rift. In contrast to the diabase which fails to provide reliable paleointensity estimates, the anorthosite xenoliths are unusually high-fidelity recorders yielding high-quality, single-slope paleointensity results that are consistent at specimen and site levels. An average value of $\sim$89 ZAm$^2$ for the virtual dipole moment from the anorthosites is higher than that of the dipole component of Earth's magnetic field today. Such a strong late Mesoproterozoic dipole recorded by the anorthosite xenoliths requires the existence of a vigorous geodynamo at the time. Together with previous paleointensity data from other Midcontinent Rift rocks, our new results indicate that a dynamo with strong power sources persisted for at least $\sim$25 million years \textit{ca.} 1.1 Ga. These data are inconsistent with there being a progressive monotonic decay of Earth's dynamo strength through the Proterozoic Eon.
\end{abstract}

@@

Paleomagnetic records from ancient rocks are one of the few types of observational data that have the potential to record the thermal evolution of Earth’s core. Evidence for a persistent magnetic field through the Proterozoic, for example, necessitates the existence of plate tectonics that sustained core-mantle boundary heat flow \cite{Swanson-Hysell2021c}. However, strikingly low estimates of geomagnetic field strengths have been obtained in the Ediacaran Period \cite{Bono2019a, Shcherbakova2019a,Thallner2021b} and the Devonian Period \cite{Shcherbakova2017a, Shcherbakova2021a}, potentially indicating unusual periods of core dynamo regime at those times. 
% These data have been interpreted to indicate that there was a progressively decaying field up to that time that was followed soon after by initial crystallization of the inner core \cite{Bono2019a}.

%\begin{figure*}
\begin{SCfigure*}
\centering
\noindent\includegraphics[width=11.4 cm]{Geologic_map.pdf}
\caption{\footnotesize{Simplified geologic map of the Lake Superior region showing the distribution of rift and post-rift-related rocks including extrusive, intrusive, and late/post-rift sediments around Lake Superior. Purple stars mark sites with paleointensity results that passed the selection criteria from this study. Paleomagnetic sites from ref. \citealp{Pesonen1983a} are categorized by lithology. All sites from ref. \citealp{Pesonen1983a, Kulakov2013a, Sprain2018a} are color-coded by their ages.}}
\label{fig:Geologic_map}
%\end{figure*}
\end{SCfigure*}

Determinations of the absolute value of the the ancient geomagnetic field strength rely on igneous rocks that acquire thermal remanent magnetizations as they cool. These magnetizations need to be unmodified by subsequent heating or chemical alteration in order to maintain a record of the ancient geomagnetic field from the time of cooling. Intracontinental magmatic events are therefore an important target for determination of ancient paleointensity as they can be well-preserved within continental interiors. This interior position results in them typically being distant from tectonic events along continental margins that can drive alteration through heat and fluids. However, large igneous provinces that are often associated with intraplate magmatism are typically of geologically short duration with the bulk of magmatic products emplaced within 1 Myr or less \cite{Kasbohm2021a}. The Midcontinent Rift is an exception as it is a large igneous province where magmatism lasted $\sim$25 Myr from \texit{ca.} 1109 Ma to 1084 Ma \cite{Swanson-Hysell2019a}. Additionally, extension ceased in the Midcontinent Rift prior to lithospheric separation, preserving volcanic, intrusive, and sedimentary rocks of the rift within the continental interior. As a result, rocks of the rift have unusually simple paleomagnetic behavior for their greater than one billion-year-old age and form a central record of Mesoproterozoic paleogeography \cite{Swanson-Hysell2021c}. The duration of magmatic activity within the Midcontinent Rift is longer than the entire 20.4 Myr long Neogene Period such that it enables an extended well-preserved window into the intensity of Earth's magnetic field in the late Mesoproterozoic.

Despite the excellent preservation of the rocks, non-ideal paleointensity behaviors and overall low success rates have often led to high failure rates and challenged the interpretation of many previous paleointensity results from the Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a}. The most trusted type of paleointensity estimate is that obtained through experiments in which the primary natural thermal remanent magnetization is progressively replaced by a laboratory magnetization that is imparted in a known field with internal consistency checks (such as in the IZZI-style Thellier experiments; ref. \citealp{Yu2004a}). In such Thellier paleointensity experiments, one typical departure from ideal behavior is sagging or double-slopes as visualized in plots that show thermal remanent magnetization (TRM) acquired versus natural remanent magnetization (NRM) lost. For such data, distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope. Typically, such non-ideal behavior would result in a higher paleointensity estimate from the steeper-sloped low-temperature portion and a lower paleointensity estimate from the high-temperature portion. For example, in data from the Midcontinent Rift, ref. \citealp{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas ref. \citealp{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Results with such non-ideal behavior were rejected by ref. \citealp{Sprain2018a} who applied stricter paleointensity selection criteria, but as a result had few accepted sites.

In this study, we target a unique rock type---high-purity anorthosite xenoliths. Anorthosites are cumulate rocks composed dominantly of the mineral plagioclase. They are attractive targets for paleomagnetic study as plagioclase crystals can enshroud minute magnetic inclusions from alteration with alteration of the plagioclase crystals not resulting in the formation of secondary iron oxides. The anorthosite xenoliths targeted in this study were brought to the near surface in magma that formed hypabyssal intrusions of the Beaver River diabase \cite{Zhang2021b}. They cooled and acquired their magnetizations in conjunction with the host diabase. Paleointensity experiments on the anorthosite xenoliths have a very high success rate, yielding consistent specimen- and site-level paleointensity results. Rock magnetic data reveal that the anorthosite xenoliths have low anisotropy of thermal remanent magnetization (TRM) and can acquire TRM linearly within relevant field strengths. Magnetic imaging shows the anorthosite specimens have dominant magnetic carriers interstitial to and within plagioclase crystals and they do not display strong preferred orientations. Our single-slope, high quality paleointensity data add to the reliable paleointensity records during $\sim$25 Myr of Midcontinent Rift magmatic activity and require that there was a strong late Mesoproterozoic geodynamo.

@@

While magmatic activity within the Midcontinent Rift was protracted, there were intervals of particularly rapid volcanism and voluminous emplacement of intrusions. The \textit{ca.} 1092 Ma Beaver Bay Complex in northern Minnesota punctuates one such period of magmatism during the main stage of Midcontinent Rift activity. The magma that formed the 1091.7 $\pm$ 0.2 Ma Beaver River diabase dikes and sills of the Beaver Bay Complex transported numerous anorthosite xenoliths that have short-axis diameters up to 180 meters via wide conduits \cite{Boerboom2004a, Boerboom2006b}. These anorthosite xenoliths are plagioclase cumulates that formed comagmatically with the host diabase in the lower crust---an interpretation confirmed by U-Pb zircon geochronology \cite{Zhang2021b}.

@@

Thermal modeling results and paleomagnetic directional data show the anorthosite xenoliths to have acquired thermal remanent magnetizations while cooling with the Beaver River diabase \cite{Zhang2021b}. Step-wise thermal demagnetization data show the anorthosite xenoliths have dominantly single-component magnetization that often unblock sharply within temperature ranges between 500\textdegree C and 580\textdegree C, consistent with them being held by low titanium titanomagnetite \cite{Zhang2021b}.

%\begin{figure*}
\begin{SCfigure*}
\centering
\noindent\includegraphics[width=10.5 cm]{Petro_QDM.png}
\caption{\footnotesize{Thin section petrographic and magnetic images of anorthosites from the Beaver River anorthosite xenolith from which paleomagnetic site AX16 and geochronology sample MS99033 were collected (left column; ref. \citealp{Zhang2021b}) and distinct anorthosite from the Duluth Complex Anorthositic Series (right column). The Duluth Complex anorthosites were not targeted for paleointensity experiments in this study given the complexities associated with more pronounced fabrics. Cross-polarized petrographic images (A) Beaver River anorthosite have plagioclase with a granoblastic texture; (B) Plagioclase of Duluth Complex anorthosite have euhedral, interlocking crystals showing igneous foliation; (C) zoomed-in image of plagioclase from MS99033 showing the crystals to be largely free of large opaque inclusions; (D) Duluth Complex plagioclase can contain abundant Fe-Ti oxide needles that have preferred orientations that are often parallel with the [010] planes of the plagioclase crystals. Magnetic field maps acquired by the quantum diamond microscope (QDM) show that the relatively weak magnetic sources within and interstitial to plagioclase crystals exist in the Beaver River anorthosite in (E) while large oxide needles within Duluth Complex plagioclase are strongly magnetic as evidenced through the overlapping magnetic fields above the oxides in (F). Magnetic field maps of (E) and (F) share the same spatial and magnetic scale. The unit of the axes are in $\mu$m.}}
\label{fig:Petro_QDM}
%\end{figure*}
\end{SCfigure*}

\subsection*{Petrography and magnetic imaging of anorthosite xenoliths}

Petrographic imaging shows the dominantly monomineralic anorthosite xenoliths within the Beaver River diabase often have granoblastic texture characterized by equigranular crystals with weakly developed petrofabrics (Fig. \ref{fig:Petro_QDM}A,C). In addition, the plagioclase crystals of the anorthosite xenoliths are typically free of large oxide inclusions. Magnetic imaging using the quantum diamond microscope (QDM) shows that the dominant magnetic remanence carriers often exist within, and sometimes interstitial to, plagioclase crystals (Fig. \ref{fig:Petro_QDM}E).

These features of the Beaver River anorthosite xenoliths are distinct from other anorthosites such as the anorthosite xenoliths within the Duluth Complex Anorthositic Series rocks---older intrusions within the Midcontinent Rift. To illustrate these difference, we present petrographic and magnetic imaging data from a sample of Duluth Complex anorthosite that was not targeted for paleointensity experiments. The plagioclase of the Duluth Complex anorthosites typically develop interlocking textures that display strong igneous foliation (Fig. \ref{fig:Petro_QDM}B). In addition, there are abundant Fe-Ti oxide needles that are typically aligned with the [010] planes of plagioclase crystals (Fig. \ref{fig:Petro_QDM}D). Magnetic imaging confirms that those strongly magnetic needles within the plagioclase crystals often have magnetic moments oriented along their long axes (Fig. \ref{fig:Petro_QDM}D,F). Titanomagnetite and ilmenite symplectic intergrowths also coexist with pyroxene and relict olivine as a product of olivine oxidation in the Duluth Complex plagioclase \cite{Miller1990a}.

The lack of symplecites and the lack of large plagioclase-hosted titanomagnetite needles within the Beaver River anorthosite xenoliths distinguish them from plagioclase cumulates of the Duluth Complex and other layered mafic complex where anisotropic magnetic mineral fabrics associated with igneous foliation often occur \cite{Scofield1986a, Selkin2000a, Feinberg2006a}. The relative lack of fabric and associated preferred orientation makes the Beaver River anorthosite xenoliths a particularly compelling target for paleointensity experiments.

@@

% Despite that the anorthosite xenoliths of the Beaver River diabase were entrained as xenoliths and emplaced only $\sim$4 Myr after the formation of the Anorthositic Series rocks of the Duluth Complex when massive plagioclase mushes were generated, $^{206}$Pb/$^{238}$U zircon geochronology data and zircon rare earth element diffusion modeling results suggest it is the most probable that the Beaver River diabase anorthosite xenoliths are entrained cumulate enclaves that formed at the time of Beaver Bay Complex magmatism. The anorthosite xenoliths of the \textit{ca.} 1092 Ma Beaver River diabase has distinct mineralogy from that of the Anorthositic Series rocks of the Duluth Complex. The dominantly monomineralic anorthosite xenoliths have an average anorthite content of $\sim$70\% \cite{Morrison1983a, Doyle2016a}, higher than the average of $\sim$60\% of the Anorthositic Series rocks of the Duluth Complex \cite{Miller1990a}. In addition, petrographic imaging support that the plagioclase from the Beaver River anorthosite formed under a magmatic conditions different from the Duluth Complex anorthositic series magma. While the plagioclase in the Duluth Complex Anorthositic Series typically are euhedral with an interlocking texture (Fig. \ref{Petro_QDM}A), the more calcium-rich plagioclase of the Beaver River anorthosite often develop granoblastic texsture characterized by equigranular plagioclase crystals (Fig. \ref{Petro_QDM}B). Abundant magnetite and ilmenite needles that are typically 10 $\mu$m long can be found to have preferred orientations within Duluth Complex plagioclase. Interstitial vermicular titanomagnetite and ilmenite symplectic intergrowths also coexist with pyroxene and relict olivine as a product of olivine oxidation in the Duluth Complex Anorthositic Series rocks \cite{Miller1990a}.

@@

Following IZZI-modified paleointensity experiments \cite{Yu2004a}, 40 from a total of 86 anorthosite specimens and 0 out of a total of 69 diabase specimens passed our paleointensity result selection criteria (see Materials and Methods section). 7 anorthosite sites and no diabase site-level results pass these selection criteria. Example NRM/TRM (Arai) plots are shown in Fig. \ref{fig:IZZI_examples}. Summary specimen absolute paleointensity estimates and site-level weighted mean paleointensity values are plotted in Fig. \ref{fig:PINT_cooling_corrected} (and provided in Table S1). The paleointensity quality index (Q$_{PI}$; ref. \citealp{Biggin2014a}) for the anorthosite xenoliths are all above 4 (Table S2). The cooling rate-corrected absolute paleointensity estimates from all anorthosite sites have a weighted mean of 52.72 $\pm $5.72 $\mu$T. This corresponds to a calculated virtual dipole moment of $\sim$89 ZAm$^2$ ($10^{21} Am^2$) \textit{ca.} 1092 Ma. All measurement-level paleointensity experiment data is available within the MagIC database (\url{}; ATTACHED IS A LINK FOR REVIEWING PURPOSESS, THE URL WILL BE UPDATED WHEN DOI IS GENERATED).

\begin{figure*}
\noindent\includegraphics[width=17.8cm]{IZZI_examples.pdf}
\centering
\caption{\footnotesize{Example results of paleointensity experiments are displayed on NRM/TRM plots and zero-field heating results are shown on inset orthogonal plots (Zijderveld plots) for anorthosite and diabase specimens. Red (blue) circles indicate zero-field/in-field (in-field/zero-field) steps `ZI’ (`IZ’). Triangles mark partial thermal remanent (pTRM) checks. Blue and red squares in the Zijderveld plots are X–Y and X–Z projections, respectively, of the NRMs in specimen coordinates. Plots on the top row show successful specimen paleointensity results with straight, single-slope behaviors that pass our selection criteria. The green lines represent fits for the dominant single-slope component that passes the acceptance criteria and gives an estimate of the ancient field strength (B$_{anc}$). The plots for anorthosite specimens AX1-1a and diabase BD5-1b on the bottom row show typical non-ideal sagging and zigzagging behaviour that fails our acceptance criteria. Specimen BD2-5a is an example where the data appear linear with distinct slopes in the low and high temperature ranges such that it could pass less restrictive selection criteria, particularly if a more narrow temperature range was used for the experiment.}}
\label{fig:IZZI_examples}
\end{figure*}

% Paleointensity experiment results from the anorthosite xenoliths and their host diabase show three general types of dominant specimen behaviors on NRM/TRM plots (Fig. \ref{fig:IZZI_examples}): dominantly single-slope, double-slope or sagging, and zigzagging. 
Typical paleointensity experimental data of the anorthosite specimens have straight, single-slope NRM/TRM plots and the accepted fractions of temperature steps span over the origin-trending, primary remanence components (Fig. \ref{fig:IZZI_examples}). We accept specimen- and site-level absolute paleointensity results from those anorthosite xenoliths that pass the selection criteria. Other anorthosite xenoliths and diabase specimens failed the selection criteria largely because of double-slope or sagging behavior (fail FRAC selection; see methods), poor pTRM checks, and sometimes zigzagging behaviors superimposed on top (fail SCAT, DANG selection; see methods; Fig. \ref{fig:IZZI_examples}). A 20 mT alternating field treatment after in-field heating steps was applied to some specimens, but it did not result in significant changes in the experimental results for the anorthosite xenolith or diabase specimens (Fig. \ref{fig:PINT_cooling_corrected}; Supporting Information).

In addition to estimating paleointensity values by introducing a set of selection criteria to filter our experiment results, we apply an independent method of ref. \citealp{Cych2021a} which uses all experimental data regardless of their NRM/TRM plot statistics to perform bias corrected estimation of paleointensity (BiCEP). This Bayesian probabilistic method is based on an assumption that paleointensity estimates from specimens that come from a same cooling unit are distributed around a true paleointensity value with the various deflections being expressed as the curvature parameter of the NRM/TRM plot \cite{Paterson2011a}. The posterior paleointensity distributions from these sites with high-quality specimen-level data are in agreement with the site-level weighted-mean results developed using the selection criteria approach (Fig. \ref{fig:PINT_cooling_corrected}; Fig. S1).

Overall, the high-quality experiment results from the anorthosite xenoliths of the Beaver River diabase and the consistent paleointensity estimates based on independent methods confirm that the anorthosite xenoliths record a high geomagnetic field \textit{ca.} 1092 Ma.

@@

\subsection*{Rock magnetic experiments}
